Dependent Object Types

### 2012-06-30 doc/semester/report.tex
first draft of patches sec

In the original DOT calculus, the \textsc{app} and \textsc{new} typing
rules have implicit relaxations. For instance, in \textsc{app}, the
argument type may be a subtypine of the declared parameter type. In
order to deal with all the soundness problems due to narrowing, we
make widening an explicit operation and changes those rules to be
strict by replacing those relaxed subtyping judgments with equality
judgments. Two types $S$ and $T$ are judged to be equal if $S <: T$
and $T <: S$.

>>>>

In the original DOT calculus, the \textsc{app} and \textsc{new} typing
rules have implicit relaxations. For instance, in \textsc{app}, the
argument type may be a subtype of the declared parameter type. In
order to deal with all the soundness problems due to narrowing, we
make widening an explicit operation and change those rules to be
strict by replacing those relaxed subtyping judgments with equality
judgments. Two types $S$ and $T$ are judged to be equal if $S <: T$
and $T <: S$.

Syntactically, we add a widening term: $\wid t T$, and extend values
with a case for widening: $\wid v T$. The typing rule for widening,
\textsc{wid}, is the only one admitting a subtyping relaxation:
$\Gamma \ts (\wid t T) \typ T$ if $\Gamma \ts t : T'$ and $\Gamma \ts T'
\sub T$.

The reduction rules become more complicated because the type
information in the widening needs to be propagated correctly. We will
motivate this informally with examples.

\begin{align*}
&\mlnew v {\Top \refine z {\Ldecl {L_a} \Bot \Top, \ldecl l {\Top \refine z {\Ldecl {L_a} \Bot \Top}}}\\
&\gap\gap\ldefs{l = \wid v {\Top \refine z {\Ldecl {L_a} \Bot \Top}}}}{
\app {(\abs x \Top x)} {(\wid v {\Top \refine z {\ldecl l \Top}}).l}
}
\end{align*}

The term $(\wid v {\Top \refine z {\ldecl l \Top}}).l$ first widens
$v$ so that the label has type $\Top$ instead of $\Top \refine z
{\Ldecl {L_a} \Bot \Top}$.

How should reduction proceed? We cannot just strip the widening and
then reduce, because then the strict function application would not
accept the reduced term. In short, we need to do some type
manipulations during reduction, by using the membership and expansion
judgments. This is a bit unfortunate, because it means that reduction
now needs to know about typing.

Next, we look at path equality provisions. These are even more
essential now in the presence of explicit widening. Consider this
example:

\begin{align*}
&\mlnew b {\Top \refine z {\Ldecl X \Top \Top, \ldecl l {z.X}} \ldefs{l = \wid b {b.X}}}{
\mlnew a {\Top \refine z {\ldecl i {\Top \refine z {\Ldecl X \Bot \Top, \ldecl l {z.X}}}}\\
&\gap\gap\ldefs{i = \wid b {\Top \refine z {\Ldecl X \Bot \Top, \ldecl l {z.X}}}}}}{
\wid {a.i.l} \Top
}
\end{align*}

$a.i.l$ reduces to $\wid b {\Top \refine z {\Ldecl X \Bot \Top, \ldecl
    l {z.X}}}$. Now, how can we continue? $b.l$ reduces to $b : b.X$
which has bounds $\Top..\Top$, but $(\wid b {\Top \refine z {\Ldecl X
    \Bot Top, \ldecl l {z.X}}}).l$ has bounds $\Bot..\Top$, so without
some provision for path equality, we cannot widen $b.l$ to $(\wid b
{\Top \refine z {\Ldecl X \Bot \Top, \ldecl l {z.X}}}).l$.

\subsection{Path Equality Provisions}

We add the path equality provisions to the subtyping rules.

Let's first ignore the extension of the calculus requiring explicit
widenings introduced in~\ref{widening}. Then, we need to add one
intuitive rule to the subtyping judgment: $\textsc{$\sub$-path}$. If
$p$ (path-)reduces to $q$, and $T \sub q.L$, then $T \sub p.L$. Path
reduction is a simplified form of reduction involving only
paths. However, this means that the subtyping judgment, and
indirectly, all the typing-related judgments, now need to carry the
store in addition to the context so that path reductions can be
calculated.

Now, let's see how path equality provisions and explicit widening can
fit together.

First, path reduction is not isomorphic to reduction anymore, since we
actually want to actually skip over widenings, as motivated by the
last example in~\ref{widening}.

In addition, we now also need a dual rule, $\textsc{path-$\sub$}$: if
$p$ (path-dually)-reduces to $q$, and $q.L \sub T$ then $p.L \sub
T$. This is because when we have a widening on an object on which a
method is called, we have to upcast the argument to the parameter type
expected by the original method. Here is a motivating example.

Let $T_c$ be a shorthand for the type:
\begin{align*}
\Top &\mlrefine z {\\
&\ \Ldecl A {\Top \refine z {\mdecl m \Bot \Top}} \Top\\
&\ \Ldecl B \Top \Top\\
&\ \mdecl m {z.A} \Top\\
}&
\end{align*}

Let $T$ be a shorthand for the type:
\begin{align*}
\Top &\mlrefine z {\\
&\ \Ldecl A {\Top \refine z {\mdecl m \Bot \Top}} \Top\\
&\ \Ldecl B \Top \Top\\
&\ \mdecl m {\Top \refine z {\Ldecl B \Top \Top}} \Top\\
}&
\end{align*}

Now, consider the term:
\begin{align*}
&\mlnew v {Tc \ldefs{m(x)=\wid x \Top}}{
(\wid v T).m(\wid v {(\wid v {A \refine z {\Ldecl B \Top \Top}})})
}
\end{align*}

When we evaluate the method invocation, we need to cast $\wid v {(\wid
  v {A \refine z {\Ldecl B \Top \Top}})}$ to $v.A$, and for this, we
need the newly introduced $\textsc{path-$\sub$}$ rule.

Note that the path dual reduction can be a bit stricter with casts
than the path reduction. In any case, introducing this
$\textsc{path-$\sub$}$ rule into the subtyping judgment is
problematic: it is now possible to say $p.L \sub T$, even though $T$
can do more than what $p.L$ is defined to do. Here is an example,
where we construct an object, with $T = \Bot$. (The convolution in the
example is due to the requirement that concrete types be only
mentioned once.)

\begin{align*}
&\mlnew a {\Top \refine z {\Ldecl C \Bot {\Top \refine z {\Ldecl D \Bot {z.X}, \Ldecl X \Bot \Top}}}}{
\mlnew b {{a.C} \refine z {\Ldecl X \Bot \Bot}}}{
\mlnew c {a.C}}{
\mlnew d {(\wid b {a.C}).D}}{
\app {(\abs x \Bot {x.\mathit{foo}})} d
}
\end{align*}

Notice that $d$ has type $\Bot$ if you ginore the cast on $b$. This
example doesn't typecheck initially because $\textsc{path-$\sub$}$
only applies when objects are in the store, so the application is not
well-typed. But if we start preservation in a store which has $a$,
$b$, $c$ and $d$ then the application type-checks, because, through
$\textsc{path-$\sub$}$, we can find that the type of $d$ is a subtype
of $\Bot$. Now, of course, when we get to $d.\mathit{foo}$, reduction
fails.

So the preservation theorem as defined on a small-step semantics
(where we start with an arbitrary well-formed environment) fails when
we add the $\textsc{path-$\sub$}$ rule.

\section{Conclucion and Future Work}
